Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-28 paper/main.tex
minor things

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 (\cite{howe2020ensembl}; Table S\ref{suptab:downloads_animals}) 
as well as all non-embargoed plant species from the JGI Phytosome 13 database 
(\cite{goodstein2012phytozome}; Table S\ref{suptab:downloads_plants}); exact 
genomes used are list here . % add a supplemental table with names and versions
Data was downloaded on 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

For both feasibility and iterative cross-species generalization validattion, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
and a variety of genome sizes and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize, if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

Once a set of training genomes was selected, we further split the sequences therein into a {\it 
training and validation set}. This split of the training genomes was done to get a 
quicker sense of the generalization capabilities and was used primarily during 
training. We split off the validation set by selecting 20\% of the FASTA sequences 
above and below the N90 of each training genome, ensuring a proper distribution of 
large and short sequences in both sets and a split on the chromosome level. 
Evaluation of the annotations on all training and validation genomes was done 
regularly after a promising model candidate was found based on its validation set 
performance. Ultimately we used the cross-species performance on all 
validation genomes as the decisive measure of model quality.

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF (https://https:
//github.com/weberlab-hhu/GeenuFF). GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

\begin{table}[!htb]
\label{tab:arrays}
\processtable{Data arrays generated and used by Helixer}{
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{{\it Note:} the encoding does not differentiate between introns
in coding and non-coding regions.}
\end{table}

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the classes of the input data are both unbalanced vary greatly 
between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. 
We compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{ssec:dcnn} of the Supplements.

>>>>

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 (\cite{howe2020ensembl}; Table S\ref{suptab:downloads_animals}) 
as well as all non-embargoed plant species from the JGI Phytosome 13 database 
(\cite{goodstein2012phytozome}; Table S\ref{suptab:downloads_plants}); The exact 
genomes are listed in the aformentioned tables.
Data was downloaded on the 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

For both feasibility and the frequent validation of cross-species generalization during
development, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs in general. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

Once a set of training genomes was selected, we further split the sequences therein into a {\it 
training and validation set}. This split of the training genomes was done to get a 
quicker sense of the generalization capabilities and was used primarily during 
training. We split off the validation set by selecting 20\% of the FASTA sequences 
above and below the N90 of each training genome at random, ensuring a proper distribution of 
large and short sequences in both sets and a split on the chromosome level. 
Evaluation of the annotations on all training and validation genomes was done 
regularly after a promising model candidate was found based on its validation set 
performance. Ultimately we used the cross-species performance on all 
validation genomes as the decisive measure of model quality.

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF 
(https://github.com/weberlab-hhu/GeenuFF). 
GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

\begin{table}[!htb]
\label{tab:arrays}
\processtable{Data arrays generated and used by Helixer}{
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{{\it Note:} The encoding does not differentiate between introns
in coding and non-coding regions.}
\end{table}

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual.

We also compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{ssec:dcnn} of the Supplements.
